Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeast Taiwan

### 2019-11-01 analysis/paper/paper.Rmd
update of code

Geometric morphometry methods (GMM) has been widely applied recently in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interests. Different from linear measurement approach that captures shapes by measuring distances, angles or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define the shapes. Landmarks, curves or outlines of objects can be transformed into coordinates by their unique location and orientation with respect to numerical values of coordinate axes [@Adams2004; @Lawing2010; @Slice2007]. According to different focuses and forms of original shape data, there are two common morphometric methods used to capture and analyze the shapes, landmark analysis and outline analysis [@Adams2004]. Landmark approaches apply landmark points to specimen as variables that can be specified on coordinate system. Based on the points, objects are translated and scaled into the same orientation and size using Procrustes superimposition [@Lawing2010]. This method commonly applied to zooarchaeology and

@@

The scanned pottery drawings of reconstructed pots that mostly covers from rim to bottom were acquired from Bureau of Cultural Affairs in Yilan. All drawing presents two-dimensional view of the section of a vessel with indications of metric measurements. The scanned drawings were imported into the Inkscape software for outlines tracing to remove some information such as marks, lines, and numbers on the original drawing. Each traced half cross-section image was duplicated, flipped, and then joined with another one to create a 2D closed outline that can be processed later in the R software (www.rproject.org, Core-Team, 2015). Geometric morphometric analyses were conducted using the functions included in the Momocs, a R package intended to quantify the shape and compare its variation, especially for outline analysis [@Bonhomme2014]. The digitised outlines were converted into a list of successsive x-y pixel coordinates for elliptic Fourier analysis (EFA), which assesses morphological differences among pottery shapes from three occupation contexts. EFA is a powerful method to extract geometric information where closed outlines are considered as periodic functions and decomposed into a series of trigonometric functions referred as harmonic coefficients [@Claude2008; @Bonhomme2014]. The number of harmonics determines the quality of the geometry that more harmonics can capture the shapes more precisely [@Bonhomme2013]. Here we used \13\ harmonics to describe pottery shapes. The harmonic coefficients generated by EFA were analysed by principal component analysis (PCA) to illustrate the diversity of the shape data and identify the major patterns of variation through dimensionality reduction. The principal components (PCs) scores were analysed with a multivariate analysis of variance (MANOVA) to test for significant effects of the groups of occupation context on shape variances. Finally, we computed a significance test to assess the equality of coefficients of variation (CVs) among multiple groups of PCs to understand the differences in shape standardisation among context groups. The R package cvequality (Version 0.2.0; Marwick and Krishnamoorthy 2019) in the CRAN repository was used to conduct the significance test for CVs.

@@

# convert svgs to to jpg
library(magick)
library(rsvg)
map(filled_svgs,
  ~image_read_svg(.x) %>% 
  image_background("white") %>%  
  image_convert("jpg") %>% 
  image_write(str_glue("{.x}_output.jpg")))
```

>>>>

Geometric morphometry methods (GMM) has been widely applied recently in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interests. Different from linear measurement approach that captures shapes by measuring distances, angles or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define the shapes. Landmarks, curves or outlines of objects can be transformed into coordinates by their unique location and orientation with respect to numerical values of coordinate axes [@Adams2004; @Lawing2010; @Slice2007]. According to different focuses and forms of original shape data, there are two common morphometric methods used to capture and analyze the shapes, landmark analysis and outline analysis [@Adams2004]. Landmark approaches apply landmark points to specimen as variables that can be specified on a coordinate system. According to the specific location of each point, objects are translated and scaled into the same orientation and size using Procrustes superimposition [@Lawing2010]. This method can be applied to 3D objects not just limited to 2D that commonly applied to zooarchaeology and

@@

The scanned pottery drawings of reconstructed pots that mostly covers from rim to bottom were acquired from Bureau of Cultural Affairs in Yilan. All drawing presents two-dimensional view of the section of a vessel with indications of metric measurements. The scanned drawings were imported into the Inkscape software for outlines tracing to remove some information such as marks, lines, and numbers on the original drawing. Each traced half cross-section image was duplicated, flipped, and then joined with another one to create a 2D closed outline that can be processed later in the R software (www.rproject.org, Core-Team, 2015). Geometric morphometric analyses were conducted using the functions included in the Momocs, a R package intended to quantify the shape and compare its variation, especially for outline analysis [@Bonhomme2014]. The digitised outlines were converted into a list of successsive x-y pixel coordinates for elliptic Fourier analysis (EFA), which assesses morphological differences among pottery shapes from three occupation contexts. EFA is a powerful method to extract geometric information where closed outlines are considered as periodic functions and decomposed into a series of trigonometric functions referred as harmonic coefficients [@Claude2008; @Bonhomme2014]. The number of harmonics determines the quality of the geometry that more harmonics can capture the shapes more precisely [@Bonhomme2013]. Here we used \13\ harmonics to describe pottery shapes.

The harmonic coefficients generated by EFA were analysed by principal component analysis (PCA) to illustrate the diversity of the shape data and identify the major patterns of variation through dimensionality reduction. The principal components (PCs) scores were analysed with a multivariate analysis of variance (MANOVA) to test for significant effects of the groups of occupation context on shape variances. Finally, we computed a significance test to assess the equality of coefficients of variation (CVs) among multiple groups of PCs to understand the differences in shape standardisation among context groups. The R package cvequality (Version 0.2.0; Marwick and Krishnamoorthy 2019) in the CRAN repository was used to conduct the significance test for CVs.

@@

# convert svgs to to jpg
library(magick)
library(rsvg)
map(filled_svgs,
  ~image_read_svg(.x) %>% 
  image_background("white") %>%  
  image_convert("jpg") %>% 
  image_write(str_glue("analysis/data/raw_data/pottery_jpg/{basename(.x)}_output.jpg")))
```
